Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-11-18 numerical-amr.tex
Very minor clean-up.

In the first case, we allow for four types of boundary conditions (for
simplicity, represented here as being at $x=0$, with positive $x$
interior and negative $x$ exterior to the domain):

\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a
    mirror, with the solution on the boundary reflecting the solution
    in the computational domain ($q(-x) = q(+x)$) but with the
    velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by
    duplicating the solution at the edge of the computational domain:
    $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a
    pre-determined function ($q(-x) = q0(-x,t)$).  The code provides a
    simple way to set inflow values that are constant in both time and
    space over an entire face, or hooks are provided for special cases
    where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the
    other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

\item{\em Baryon mass.}  This refinement criterion is designed to
  mimic a Lagrangian method in that it tries to keep a fixed mass
  resolution.  In this method, a cell is refined if the baryonic mass
  in the cell ($M_g = \rho\, (\Delta x)^d$) is larger than a specified value:

\begin{equation}
M_g > \rho_{\rm flag}\, (\Delta x_{\rm root})^d r^{\epsilon_l l},
\end{equation}

where $\rho_{\rm flag}$ is the equivalent density on the root grid
required for refinement, $\Delta x_{\rm root}$ is the root grid cell
spacing, $r$ is the refinement factor, $l$ is the level and
$\epsilon_l$ is a parameter that can be used to make the refinement
more aggressive ($\epsilon_l < 0$) or less aggressive ($\epsilon_l >
0$) than pure Lagrangian refinement --- in other words,
super-Lagrangian or sub-Lagrangian refinement.

\item{\em Shocks.}  In this method, we identify and refine cells that
  contain shocks using the following criteria:

\begin{eqnarray}
(p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) > 0.33,  \qquad
u_{i-1} - u_{i+1} > 0, \quad\text{and} \qquad
e_i / E_i > 0.1.  \nonumber
\end{eqnarray}

We note that the parameters $0.33$ and $0.1$ in the equations above
were determined empirically, and can be modified by the user to refine
more or less aggressively, as the situation warrants.

@@

\item{\em Jeans length.}  Following \cite{Truelove98}, the code can be
  forced to refine the Jeans length by a fixed number of cells, or,
  more specifically, whenever the following criterion is not met:

\begin{equation}
\Delta x < \left( \frac{\gamma \pi k_B T}{N_J^2 G \rho \mu \mh} \right)^{1/2},
\end{equation}
where $N_J$ is the required number of cells per Jeans length (4 by default).

@@

\item{\em Shear.} Originally used in \citet{Kritsuk06} and later 
  modified into a more general form, turbulence calculations and
  Kelvin-Helmholtz instabilities can be successfully modeled with AMR if 
  the shear is used as a refinement parameter; in particular, refinement 
  occurs when the finite difference version of the following inequality
  holds:

\begin{equation}
\left| \sum_{i \ne j} \left( \frac{\partial v_i}{\partial x_j} + \frac{\partial v_j}{\partial x_i} \right) \right|
> \epsilon_s \frac{c_s}{\Delta x},
\end{equation}
where $\epsilon_s$ is a dimensionless parameter.

\item{\em Optical depth.} This refinement criterion, typically used
for simulations using radiation transport, flags cells for refinement
if they have an \ion{H}{1} optical depth lower than unity.  The
optical depth of a cell is estimated to be:

\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}

where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same
along each axis).  If $\tau > 1$, the cell is refined further, up to
the maximum level of refinement.

\item{\em Resistive length.}  In a way similar to the Jeans length
  criterion described above, the code can be forced to locally resolve
  the resistive length scale by a minimum of a fixed number of cells.
  This is done by ensuring that the resistive length scale, defined as

\begin{equation}
L_R = |B| / |\curl B|
\end{equation}

is always refined by at least N$_R$ cells locally (i.e.~we refine a
given cell further if $L_R \leq N_R \Delta x$).  Note that \enzo\ does
not actually solve the resistive MHD equations -- this is the
``characteristic'' local resistive length, rather than one calculated
within the MHD solver.

@@

\item{\em Second Derivative.}  This is similar to the refinement criteria used
  in Paramesh and was originally developed by \cite{1987nasa.reptQ....L}.  This
  method calculates the second derivative of a user-defined field normalized by
  the first derivative, and sets a threshold for refinement.  More background
  information can be found in the FLASH4 user guide\footnote{\url{http://flash.uchicago.edu/site/flashcode/user\_support/flash4\_ug/}},
  but we reproduce the relevant equation here for completeness. The
  multi-dimensional version of the criteria takes in to account all
  cross-derivatives, and is expressed by

 \begin{equation}
   S = \left\{ \frac{\sum_{ij}\left( \frac{\partial^2f}{\partial x_i \partial x_j} \right)^2}{\sum_{ij}\left[ \frac{1}{2\Delta x_j}\left( \abs{\frac{\partial f}{\partial x_i}}_{x_i+\Delta x_i} + \abs{\frac{\partial f}{\partial x_i}}_{x_i-\Delta x_i} \right) + \epsilon \frac{\abs{\bar{f_{ij}}}}{\Delta x_i \Delta x_j} \right]^2} \right\}^{1/2},
   \label{eq:second-deriv_analytical}
 \end{equation}

  where $f$ is the user-defined field, $\abs{\bar{f_{ij}}}$ is an average of
  $f$ over cells in the $i$ and $j$ dimensions, and $\epsilon$ is
  a user-controlled value that provides a method of ignoring small
  fluctuations, set by default to $10^{-2}$.  If $S$ is larger than a specified
  threshold (bounded by [0,1]) a cell is marked for refinement. This threshold
  is chosen at runtime and can vary for each of the fields under consideration.

>>>>

In the first case, we allow for four types of boundary conditions (for
simplicity, represented here as being at $x=0$, with positive $x$
interior and negative $x$ exterior to the domain):
\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a
    mirror, with the solution on the boundary reflecting the solution
    in the computational domain ($q(-x) = q(+x)$) but with the
    velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by
    duplicating the solution at the edge of the computational domain:
    $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a
    pre-determined function ($q(-x) = q0(-x,t)$).  The code provides a
    simple way to set inflow values that are constant in both time and
    space over an entire face, or hooks are provided for special cases
    where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the
    other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

\item{\em Baryon mass.}  This refinement criterion is designed to
  mimic a Lagrangian method in that it tries to keep a fixed mass
  resolution.  In this method, a cell is refined if the baryonic mass
  in the cell ($M_g = \rho\, (\Delta x)^d$) is larger than a specified value:
\begin{equation}
M_g > \rho_{\rm flag}\, (\Delta x_{\rm root})^d r^{\epsilon_l l},
\end{equation}
where $\rho_{\rm flag}$ is the equivalent density on the root grid
required for refinement, $\Delta x_{\rm root}$ is the root grid cell
spacing, $r$ is the refinement factor, $l$ is the level and
$\epsilon_l$ is a parameter that can be used to make the refinement
more aggressive ($\epsilon_l < 0$) or less aggressive ($\epsilon_l >
0$) than pure Lagrangian refinement --- in other words,
super-Lagrangian or sub-Lagrangian refinement.

\item{\em Shocks.}  In this method, we identify and refine cells that
  contain shocks using the following criteria:
\begin{eqnarray}
(p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) > 0.33,  \qquad
u_{i-1} - u_{i+1} > 0, \quad\text{and} \qquad
e_i / E_i > 0.1.  \nonumber
\end{eqnarray}
We note that the parameters $0.33$ and $0.1$ in the equations above
were determined empirically, and can be modified by the user to refine
more or less aggressively, as the situation warrants.

@@

\item{\em Jeans length.}  Following \cite{Truelove98}, the code can be
  forced to refine the Jeans length by a fixed number of cells, or,
  more specifically, whenever the following criterion is not met:
\begin{equation}
\Delta x < \left( \frac{\gamma \pi k_B T}{N_J^2 G \rho \mu \mh} \right)^{1/2},
\end{equation}
where $N_J$ is the required number of cells per Jeans length (4 by default).

@@

\item{\em Shear.} Originally used in \citet{Kritsuk06} and later 
  modified into a more general form, turbulence calculations and
  Kelvin-Helmholtz instabilities can be successfully modeled with AMR if 
  the shear is used as a refinement parameter; in particular, refinement 
  occurs when the finite difference version of the following inequality
  holds:
\begin{equation}
\left( \frac{\partial v_x}{\partial y} + \frac{\partial v_y}{\partial x} \right)^2 +
\left( \frac{\partial v_z}{\partial y} + \frac{\partial v_y}{\partial z} \right)^2 +
\left( \frac{\partial v_x}{\partial z} + \frac{\partial v_z}{\partial x} \right)^2
 >  \left(\epsilon_s \frac{c_s}{\Delta x}\right)^2
\end{equation}
where $\epsilon_s$ is a dimensionless user-defined parameter and $\Delta x$ 
is the local cell size at the level where refinement is occurring.

\item{\em Optical depth.} This refinement criterion, typically used
for simulations using radiation transport, flags cells for refinement
if they have an \ion{H}{1} optical depth lower than unity.  The
optical depth of a cell is estimated to be:
\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}
where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same
along each axis).  If $\tau > 1$, the cell is refined further, up to
the maximum level of refinement.

\item{\em Resistive length.}  In a way similar to the Jeans length
  criterion described above, the code can be forced to locally resolve
  the resistive length scale by a minimum of a fixed number of cells.
  This is done by ensuring that the resistive length scale, defined as
\begin{equation}
L_R = |B| / |\curl B|
\end{equation}
is always refined by at least N$_R$ cells locally (i.e.~we refine a
given cell further if $L_R \leq N_R \Delta x$).  Note that \enzo\ does
not actually solve the resistive MHD equations -- this is the
``characteristic'' local resistive length, rather than one calculated
within the MHD solver.

@@

\item{\em Second Derivative.}  This is similar to the refinement criteria used
  in Paramesh and was originally developed by \cite{1987nasa.reptQ....L}.  This
  method calculates the second derivative of a user-defined field normalized by
  the first derivative, and sets a threshold for refinement.  More background
  information can be found in the FLASH4 user guide\footnote{\url{http://flash.uchicago.edu/site/flashcode/user\_support/flash4\_ug/}},
  but we reproduce the relevant equation here for completeness. The
  multi-dimensional version of the criteria takes in to account all
  cross-derivatives, and is expressed by,
 \begin{equation}
   S = \left\{ \frac{\sum_{ij}\left( \frac{\partial^2f}{\partial x_i \partial x_j} \right)^2}{\sum_{ij}\left[ \frac{1}{2\Delta x_j}\left( \abs{\frac{\partial f}{\partial x_i}}_{x_i+\Delta x_i} + \abs{\frac{\partial f}{\partial x_i}}_{x_i-\Delta x_i} \right) + \epsilon \frac{\abs{\bar{f_{ij}}}}{\Delta x_i \Delta x_j} \right]^2} \right\}^{1/2},
   \label{eq:second-deriv_analytical}
 \end{equation}
  where $f$ is the user-defined field, $\abs{\bar{f_{ij}}}$ is an average of
  $f$ over cells in the $i$ and $j$ dimensions, and $\epsilon$ is
  a user-controlled value that provides a method of ignoring small
  fluctuations, set by default to $10^{-2}$.  If $S$ is larger than a specified
  threshold (bounded by [0,1]) a cell is marked for refinement. This threshold
  is chosen at runtime and can vary for each of the fields under consideration.
